---
title: "COVID-19 and the Big 5 Personality Test"
format: gfm
editor: visual
---

#### A Replication of <https://tylerburleigh.com/blog/covid-19-and-the-big-5-personality-test/>

##### by Lital, Intithar, and Louis

Tyler Burleigh poses the question: Does the growth in COVID-19 cases have anything to do with Big 5 Personality traits?

Here we replicate his work in R, fixing some minor mistakes and adding a bonus visualization.

```{r}
library(tidyverse)
library(lubridate)
library(ggpubr) # for easy plotting later

# load preprocessed data
load("processed_data.RData")
```

## COVID Data Preprocessing

Since the raw data is too large to upload to Github, the code here is just for display. The final preprocessed data can be found in `processed_data.RData`.

```{r}
#| eval: FALSE

# Import Data
covid <- read_csv("train.csv")

# Select the columns we need
covid <- covid %>% 
  arrange(`Country/Region`, Date) %>% 
  rename(country = `Country/Region`, date = Date, cases = ConfirmedCases) %>% 
  group_by(country, date) %>% 
  summarise(cases = sum(cases)) %>% 
  select(country, date, cases)

covid <- covid %>% 
  # Filter only countries with at least 50 cases
  # (this will also cut out all days before arriving at 50 cases)
  group_by(country) %>% 
  filter(cummax(cases) >= 50) %>% 
  # ...and at least 14 days of data afterwards 
  # (this will happen automatically when a country has no value for day 14)
  arrange(country, date) %>% 
  mutate(day = 1:n()) %>%
  filter(day %in% c(1, 14)) %>% 
  pivot_wider(id_cols = 'country', names_from = 'day', values_from = 'cases', names_prefix = "cases_day") %>% 
  # compute growth over 14 days
  mutate(growth = cases_day14 - cases_day1) %>% 
  drop_na()
```

## Big 5 Data Preprocessing

```{r}
#| eval: FALSE
# Note: filepath should be changed as needed
bigfive <- read_tsv("bigfive_data.csv")
bigfive <- bigfive %>% select(EXT1:OPN10, country) %>% 
  mutate(across(-country, as.numeric))

negatively_keyed = c('EXT2', 'EXT4', 'EXT6', 'EXT8', 'EXT10',
                    'EST2', 'EST4',
                    'AGR1', 'AGR3', 'AGR5', 'AGR7', 
                    'CSN2', 'CSN4', 'CSN6', 'CSN8', 
                    'OPN2', 'OPN4', 'OPN6')

# WARNING: this takes a long time to run
bigfive <- bigfive %>% 
  # reverse negatively keyed variables
  mutate(across(negatively_keyed, ~(6-.x))) %>% 
  # average each trait within individual
  mutate(EXT = rowMeans(pick(EXT1:EXT10), na.rm = TRUE),
         EST = rowMeans(pick(EST1:EST10), na.rm = TRUE),
         AGR = rowMeans(pick(AGR1:AGR10), na.rm = TRUE),
         CSN = rowMeans(pick(CSN1:CSN10), na.rm = TRUE),
         OPN = rowMeans(pick(OPN1:OPN10), na.rm = TRUE)) %>% 
  select(EXT:OPN, country)

bigfive <- bigfive %>% 
  # average by country, noting how many cases are in each
    # (we'll keep countries with 1000 or fewer cases for now, 
    # just in case we want them for visualization down the line)
  # just for kicks, lets take note of standard deviation too
  group_by(country) %>% 
  summarise(EXT_sd = sd(EXT, na.rm = TRUE),
            EST_sd = sd(EST, na.rm = TRUE),
            AGR_sd = sd(AGR, na.rm = TRUE),
            CSN_sd = sd(CSN, na.rm = TRUE),
            OPN_sd = sd(OPN, na.rm = TRUE),
            
            EXT = mean(EXT, na.rm = TRUE),
            EST = mean(EST, na.rm = TRUE),
            AGR = mean(AGR, na.rm = TRUE),
            CSN = mean(CSN, na.rm = TRUE),
            OPN = mean(OPN, na.rm = TRUE),
            
            n = n()) 

# Import country codes
country_codes <- read_csv("country_codes.csv") %>% select(1:2) %>% 
  rename(country = `English short name lower case`,
         code = `Alpha-2 code`)

# Put all the data together
unified_data <- bigfive %>% 
  rename(code = country) %>% 
  left_join(country_codes) %>% 
  # I'm noticing that not all names match up, so we'll need to change a couple manually before the big join
  mutate(country = case_when(country == 'United States Of America' ~ 'US',
                             country == 'Korea, Republic of (South Korea)' ~ 'Korea, South',
                             .default = country)) %>% 
  right_join(covid)
```

## Original Plots

```{r}
EXT_PLOT <- ggscatter(unified_data, x = "EXT", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Extraversion ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "EXT", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
EXT_PLOT

EST_PLOT <- ggscatter(unified_data, x = "EST", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Emotional Stability ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "EST", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
EST_PLOT

AGR_PLOT <- ggscatter(unified_data, x = "AGR", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Agreeableness ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "AGR", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
AGR_PLOT

CSN_PLOT <- ggscatter(unified_data, x = "CSN", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Conscientiousness ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "CSN", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
CSN_PLOT

OPN_PLOT <- ggscatter(unified_data, x = "OPN", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Openness ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "OPN", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
OPN_PLOT
```

## Original Plots, Now With Outliers Removed

```{r}
# Lets drop China and USA (outliers) and plot again
unified_data_without_china_usa <- unified_data %>% filter(!country %in% c('US', 'China'))

EXT_PLOT <- ggscatter(unified_data_without_china_usa, x = "EXT", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Extraversion ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "EXT", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
EXT_PLOT

EST_PLOT <- ggscatter(unified_data_without_china_usa, x = "EST", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Emotional Stability ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "EST", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
EST_PLOT

AGR_PLOT <- ggscatter(unified_data_without_china_usa, x = "AGR", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Agreeableness ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "AGR", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
AGR_PLOT

CSN_PLOT <- ggscatter(unified_data_without_china_usa, x = "CSN", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Conscientiousness ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "CSN", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
CSN_PLOT

OPN_PLOT <- ggscatter(unified_data_without_china_usa, x = "OPN", y = "growth", 
                      title = "Confirmed cases at 14 days after first 50 cases \nby average score on Big 5 factor Openness ", color = "steelblue",
                      add = "reg.line", conf.int = TRUE, 
                      cor.coef = TRUE, cor.method = "pearson",
                      xlab = "OPN", ylab = "ConfirmedCases",
                      add.params = list(color = "steelblue", fill = "lightblue"))
OPN_PLOT
```

## Bonus Plots!

```{r}
#| fig.width: 7.5
#| fig.height: 6

options(scipen = 999)
library(hrbrthemes)
library(viridis)
library(showtext)
library(sysfonts)
library(extrafont)

#------------------Extraversion--------------
  ggplot(unified_data ,aes(x= EXT, y= growth, 
                             size= n, fill=country)) +
  geom_point(alpha=0.5, shape=21, color="black") +
  scale_size(range = c(.9, 40), name="country's Cases",
                labels = scales::comma, guide = FALSE) +
 scale_fill_viridis(discrete=TRUE, guide=FALSE, option="B") +
ggrepel::geom_text_repel(aes(label = country), size = 2) +
scale_x_log10(label = scales::comma,expand = expansion(mult = c(0, .3)))+
scale_y_log10(label = scales::comma,expand = expansion(mult = c(0, .3))) +
 
theme_ipsum() +
theme(text=element_text(family="notosans"),
 legend.position="bottom") +
  labs(title="Does Extraversion Scores Predict The Growth\n in COVID-19 Cases?",
   x="Mean Extraversion", y="Increase in Cases Over 14 Days")+

ggthemes::theme_economist() +
theme(legend.text = element_text(size=10),
 plot.title = element_text(size=23),
axis.title.x = element_text(vjust = -0.8),
axis.title.y = element_text(vjust = 0.8),
 plot.caption = element_text(vjust = -1,size=9),
legend.position = c(0.4,0.85))

 #------------------Emotional Stability--------------
 ggplot(unified_data ,aes(x= EST, y= growth, 
                          size= n, fill=country)) +
   geom_point(alpha=0.5, shape=21, color="black") +
   scale_size(range = c(.9, 40), name="country's Cases",
              labels = scales::comma, guide = FALSE) +
   scale_fill_viridis(discrete=TRUE, guide=FALSE, option="B") +
   ggrepel::geom_text_repel(aes(label = country), size = 2) +
   scale_x_log10(label = scales::comma,expand = expansion(mult = c(0, .3)))+
   scale_y_log10(label = scales::comma,expand = expansion(mult = c(0, .3))) +
   
   theme_ipsum() +
   theme(text=element_text(family="notosans"),
         legend.position="bottom") +
   labs(title="Does Emotional Stability Scores Predict The Growth\n in COVID-19 Cases?",
        x="Mean Emotional Stability", y="Increase in Cases Over 14 Days")+
   
   ggthemes::theme_economist() +
   theme(legend.text = element_text(size=10),
         plot.title = element_text(size=23),
         axis.title.x = element_text(vjust = -0.8),
         axis.title.y = element_text(vjust = 0.8),
         plot.caption = element_text(vjust = -1,size=9),
         legend.position = c(0.4,0.85))
 
 #------------------Agreeableness--------------
 ggplot(unified_data ,aes(x= AGR, y= growth, 
                          size= n, fill=country)) +
   geom_point(alpha=0.5, shape=21, color="black") +
   scale_size(range = c(.9, 40), name="country's Cases",
              labels = scales::comma, guide = FALSE) +
   scale_fill_viridis(discrete=TRUE, guide=FALSE, option="B") +
   ggrepel::geom_text_repel(aes(label = country), size = 2) +
   scale_x_log10(label = scales::comma,expand = expansion(mult = c(0, .3)))+
   scale_y_log10(label = scales::comma,expand = expansion(mult = c(0, .3))) +
   
   theme_ipsum() +
   theme(text=element_text(family="notosans"),
         legend.position="bottom") +
   labs(title="Does Agreeableness Scores Predict The Growth\n in COVID-19 Cases?",
        x="Mean Agreeableness", y="Increase in Cases Over 14 Days")+
   
   ggthemes::theme_economist() +
   theme(legend.text = element_text(size=10),
         plot.title = element_text(size=23),
         axis.title.x = element_text(vjust = -0.8),
         axis.title.y = element_text(vjust = 0.8),
         plot.caption = element_text(vjust = -1,size=9),
         legend.position = c(0.4,0.85))
 
 #------------------Conscientiousness--------------
 ggplot(unified_data ,aes(x= CSN, y= growth, 
                          size= n, fill=country)) +
   geom_point(alpha=0.5, shape=21, color="black") +
   scale_size(range = c(.9, 40), name="country's Cases",
              labels = scales::comma, guide = FALSE) +
   scale_fill_viridis(discrete=TRUE, guide=FALSE, option="B") +
   ggrepel::geom_text_repel(aes(label = country), size = 2) +
   scale_x_log10(label = scales::comma,expand = expansion(mult = c(0, .3)))+
   scale_y_log10(label = scales::comma,expand = expansion(mult = c(0, .3))) +
   
   theme_ipsum() +
   theme(text=element_text(family="notosans"),
         legend.position="bottom") +
   labs(title="Does Conscientiousness Scores Predict The Growth\n in COVID-19 Cases?",
        x="Mean Conscientiousness", y="Increase in Cases Over 14 Days")+
   
   ggthemes::theme_economist() +
   theme(legend.text = element_text(size=10),
         plot.title = element_text(size=23),
         axis.title.x = element_text(vjust = -0.8),
         axis.title.y = element_text(vjust = 0.8),
         plot.caption = element_text(vjust = -1,size=9),
         legend.position = c(0.4,0.85))
 
 #------------------Openness--------------
 ggplot(unified_data ,aes(x= OPN, y= growth, 
                          size= n, fill=country)) +
   geom_point(alpha=0.5, shape=21, color="black") +
   scale_size(range = c(.9, 40), name="country's Cases",
              labels = scales::comma, guide = FALSE) +
   scale_fill_viridis(discrete=TRUE, guide=FALSE, option="B") +
   ggrepel::geom_text_repel(aes(label = country), size = 2) +
   scale_x_log10(label = scales::comma,expand = expansion(mult = c(0, .3)))+
   scale_y_log10(label = scales::comma,expand = expansion(mult = c(0, .3))) +
   
   theme_ipsum() +
   theme(text=element_text(family="notosans"),
         legend.position="bottom") +
   labs(title="Does Openness Scores Predict The Growth\n in COVID-19 Cases?",
        x="Mean Openness", y="Increase in Cases Over 14 Days")+
   
   ggthemes::theme_economist() +
   theme(legend.text = element_text(size=10),
         plot.title = element_text(size=23),
         axis.title.x = element_text(vjust = -0.8),
         axis.title.y = element_text(vjust = 0.8),
         plot.caption = element_text(vjust = -1,size=9),
         legend.position = c(0.4,0.85))
```
